---
title: "Proj"
output: pdf_document
---

FWI- The scales start at zero, and except for the Fine Fuel Moisture Code which has a maximum of 99, all are open-ended. Low ratings indicate high moisture content, and ratings rise as moisture content decreases. Ratings rise as fire weather becomes more severe.

FFMC-This is a numerical rating of the moisture content of surface litter and other cured fine fuels. It shows the relative ease of ignition and flammability of fine fuels. The moisture content of fine fuels is very sensitive to the weather. Even a day of rain, or of fine and windy weather, will significantly affect the FFMC rating. The system uses a time lag of two-thirds of a day to accurately measure the moisture content in fine fuels. The FFMC rating is on a scale of 0 to 99. Any figure above 70 is high, and above 90 is extreme.


```{r}
getwd()
df<- read.csv("forestfires.csv")
str(df)

```

```{r}
library(ggplot2)
ggplot(df, aes(wind))  + geom_s(aes(y = FFMC, colour = "FFMC"))
plot(df$wind,df$FFMC, col = 'blue',pch = 16,xlab = 'wind', ylab = 'FFMC')
```