---
title: "Homework 1 (BA)"
output: github_document
author: "Nitish Dabas (nd1292), Harsh Yadav (hy1217), Kush Shah (ks4437)"
date: "September 21, 2017"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
                        
## 1. Load the given data “contribution.csv”
```{r}
df <- read.csv("C:/Users/Harsh Yadav/Desktop/contribution.csv", header=TRUE,stringsAsFactors=FALSE)
```

## 2.How many observations and variables are there in this dataset?
```{r}
  dim(df)
```
## Ans. 1230 Observations of 11 Variables.

## 3.Are there any missing values and anomaly values in each of the variable in the dataset?
```{r}
colSums(is.na(df))
```
## Ans. No

## 4.What year(s) does the dataset cover?
```{r}
unique(df$Class.Year)
```
## Ans. 1957 1967 1977 1987 1997
\newpage
## 5.Which year has the most alumni making contribution?
```{r}
df.y0 <- subset(df, FY00Giving >0)
nrow(df.y0) 
df.y1 <- subset(df, FY01Giving >0)
nrow(df.y1)
df.y2 <- subset(df, FY02Giving >0)
nrow(df.y2)
df.y3 <- subset(df, FY03Giving >0)
nrow(df.y3)
df.y4 <- subset(df, FY04Giving >0)
nrow(df.y4)
```
## Ans. Year FY01Giving (Calculated the number of almunis contributing each year by excluding '0' contributions and then selecting the maximum number of alumnis)

## 6.Which year has the largest average contribution?
```{r}
df.m_y0 <- mean(df$FY00Giving)
df.m_y0        #Calculated the mean contribution of all years to select the largest value
df.m_y1 <- mean(df$FY01Giving)
df.m_y1
df.m_y2 <- mean(df$FY02Giving)
df.m_y2
df.m_y3 <- mean(df$FY03Giving)
df.m_y3
df.m_y4 <- mean(df$FY04Giving)
df.m_y4
```
## Ans. FY01Givings
## 7.Which year has the largest 90th percentile contribution?
```{r}
df.p_y0 <- quantile(df$FY00Giving, probs=0.9)
df.p_y0              #Calculated 90th percentile of each year to select the largest value
df.p_y1 <- quantile(df$FY01Giving, probs=0.9)
df.p_y1
df.p_y2 <- quantile(df$FY02Giving, probs=0.9)
df.p_y2
df.p_y3 <- quantile(df$FY03Giving, probs=0.9)
df.p_y3
df.p_y4 <- quantile(df$FY04Giving, probs=0.9)
df.p_y4
```
## Ans. FY01Giving

## 8.What are the standard deviation of the contribution amount in each of this years?
```{r}
df.sd_y0 <- sd(df$FY00Giving) 
df.sd_y0                    #Calculated and printed the std of eaach other
df.sd_y1 <- sd(df$FY01Giving)
df.sd_y1
df.sd_y2 <- sd(df$FY02Giving)
df.sd_y2
df.sd_y3 <- sd(df$FY03Giving)
df.sd_y3
df.sd_y4 <- sd(df$FY04Giving)
df.sd_y4
```
## Ans. FY00Giving - 1170.638
        FY01Giving - 4663.706
        FY02Giving - 638.1267
        FY03Giving - 2062.392
        FY04Giving - 792.0376

## 9.Which degree contributed the most in the latest year in the data?
```{r}
degreelist<-c(unique(df$Next.Degree)) #Creates a list of all Degrees
j=0;                                  #Initialize a variable to 0
sum_total <- c()                      # Initialize an empty list  
for (degreename in degreelist) #Loop creates a list of total contibution of all degrees
{                               
  j=j+1;
  sum_total[j] <- (sum(subset(df$FY04Giving, df$Next.Degree==degreename)))
}
degreedata <- data.frame(degreelist, sum_total) #Created a new data frame
print (subset(degreedata$degreelist , sum_total==max(degreedata$sum_total)))
#Prints the degree with most contribution 
```
## Ans. "JD" (This result can also be deduced by printing out the total contribution of each degree along with its name and then manually searching for most contribution)
\newpage
## 10.Which major contributed the most in the latest year in the data?
```{r}
majorlist<-c(unique(df$Major)) #Creates a list of all Majors 
i=0;                           #Initialize a variable to 0
sum_total2 <- c()               # Initialize an empty list  
for (majorname in majorlist)   #Loop creates a list of total contibution of all majors
{
  i=i+1;
  sum_total2[i] <- (sum(subset(df$FY04Giving, df$Major==majorname)))
}
majordata <- data.frame(majorlist, sum_total2) #Created a new data frames 
#print(majordata)
newdata <- majordata[order(-sum_total2),] 
#newdata
newdata1<-head(newdata,10)
newdata1
print (subset(majordata$majorlist , sum_total2==max(majordata$sum_total2))) 
#Prints the major with most contribution 

library(ggplot2)

df.new<-subset(df,df$Major=="Biology"| df$Major=="History" |df$Major=="Political Science" |df$Major=="Speech (Drama, etc.)" |df$Major=="Psychology")
#df.new
ggplot(df.new, aes(x=FY04Giving, fill=Major)) +
    geom_histogram(alpha=0.5, position="identity") +
    ggtitle("Contribution in 2004")
    
```
## Ans. "History" (This result can also be deduced by printing out the total contribution of each major along with its name and then manually searching for highest contribution)
    




















